---
title: "Why Dichotomizing Continuous Variables Is a Mistake"
author: Fei Zuo
date: 2026-02-11
#date-modified: 2026-02-11
subtitle: "A small simulation showing the information loss from categorizing predictors."
categories: [regression, modelling, simulation]
---

Many applied papers still dichotomize continuous predictors. This feels simple. It is not harmless.

Let’s simulate a clean linear relationship and see what happens.

## Simulation Setup

```{r}
#| message: false
#| warning: false

set.seed(1)

n <- 500
x <- rnorm(n)
y <- 2*x + rnorm(n)

d <- data.frame(x, y)
```

## Correct Model (Continuous Predictor)

```{r}
#| code-fold: true
fit_cont <- lm(y ~ x, data = d)
summary(fit_cont)
```

## Dichotomized Model

```{r}
#| code-fold: true

d$x_cat <- ifelse(d$x > median(d$x), 1, 0)

fit_cat <- lm(y ~ x_cat, data = d)
summary(fit_cat)
```

## Comparing R-squared

```{r}
#| code-fold: true

summary(fit_cont)$r.squared
summary(fit_cat)$r.squared
```

The dichotomized model throws away information. The estimated relationship becomes less precise and less powerful.

## Takeaway

Categorizing a continuous predictor reduces statistical efficiency and can distort inference. Unless there is a compelling clinical or decision-based reason, keep predictors continuous.
